---
title: "DSinR - Chapter 7"
author: "Jan Hynek"
date: "4 listopadu 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(corrr)
```
## Exercises 7.3.4
```{r}
ggplot(diamonds) +
  geom_histogram(mapping = aes(x = x), binwidth = 0.5, fill = "blue")
```
```{r}
ggplot(diamonds) +
  geom_histogram(mapping = aes(x = z), binwidth = 0.5)

```

```{r}
diamonds %>%
  filter(z < 1 | z > 10) %>%
  select(price, x, y, z, carat)
```

```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = price), binwidth = 50)
```
We can observe interesting dent around 1500. It is also worth noting that there is more diamonds with price 4xxx rather than with price 3xxx, what is kind of counterintuitive.
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = price), binwidth = 10) + coord_cartesian(xlim = c(1000, 2000))
```
```{r}
diamonds %>%
  filter(between(carat, 0.99, 1.0)) %>%
  group_by(carat) %>%
  summarise(count = n())
```
### 7.5.1.1 Exercises

```{r}
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(x = sched_dep_time, y = ..density..)) + 
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1)

```

_What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?_

Let's have a look. Let's take all floating variables and correlate them:

```{r}
glimpse(diamonds)
diamonds %>%
  select(carat, depth, table, price, x, y, z) %>%
  correlate() %>% focus(price)
```
We can see that carats are strongly correlated with 


```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = carat)) +
  geom_boxplot()

```

We can observe, that Fair diamonds have slightly more carats the Good or Very Good cut diamonds.

```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 0.2)
```

We can observe that fair cut diamonds tend to have more carats, and this might be the reason why they are more expensive.

```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) + 
  geom_violin(mapping = aes(colour = cut))

```

### 7.5.2.1 Exercises

_How could you rescale the count dataset above to more clearly show the distribution of cut within colour, or colour within cut?_
I decided to use percentile rank of each count. This scales shows more of the variability in the data.

```{r}
diamonds %>% 
  count(color, cut) %>% mutate(p_rank = percent_rank(n)) %>%  
  ggplot(mapping = aes(x = color, y = cut)) +
    geom_tile(mapping = aes(fill = p_rank))



```

```{r}
glimpse(nycflights13::flights)
```

```{r}
nycflights13::flights %>% count(month, dest) %>%
  ggplot(mapping = aes(x = month, y = dest)) +
  geom_tile(aes(fill = n))
```


### 7.5.3.1 Exercises

_Instead of summarising the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using cut_width() vs cut_number()? How does that impact a visualisation of the 2d distribution of carat and price?_

```{r}

diamonds %>% mutate(price_perc = factor(ntile(price, 5))) %>%
  ggplot(mapping = aes(x = carat, y = ..density..)) + 
    geom_freqpoly(mapping = aes(colour = price_perc), binwidth = 0.1)
```
```{r}
diamonds %>% mutate(price_perc = factor(ntile(price, 6)))
```

