---
title: "Individual Assignment 1"
author: "Jan Hynek"
date: "19 listopadu 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading Libraries
```{r libraries, warning=FALSE,  message=FALSE}
library(tidyverse)
library(reshape2)
library(purrr)
library(viridis)
library(ggthemes)
library(assertthat)
library(plotly)
```


## Exercise 1 - Transforming X using 9 Gaussian basis functions.
_Transform the $x$ by using 9 Gaussian basis functions. Fix the parameters of your gaussian functions. Take note of the domain of the function when you select the parameter governing the spatial scale (i.e. $s$). Share your code and graph the results. You need to use your result in the succeeding questions. (1 point)_

In this part I create functions, which are used throughout the script. They will be handy, as we are creating several different sizes of the dataset.

This function creates data of the specified length. Returns two vectors in a dataframe - x, and target.
```{r data_gen}
data_gen <- function(n){
  x <- runif(n, min = -1, max = 1)
  noise <- rnorm(n, mean = 0, sd = 0.1)
  target <- sin(2 * pi * x) + noise
  return(data.frame(x, target))
}
```


This is the gaussian basis function. Transforms the vector using gaussian function $\phi_j(x) = exp\left(\frac{(x-\mu_j)^2}{2s^2}\right) $.
```{r gauss}
gauss_basis <- function(x, mu, s_sq){
  return(
    exp(
      (-(x - mu) ^ 2) / (2 * s_sq)
    )
  )
}
```

This function takes dataset created from data_gen, and transforms it using 9 Gaussian basis functions. returns n times 11  matrix, where 1st columns is x, 2nd is target and next 9 are transformed gaussian bases. I can also modify $s^2$, however it is preset to 0.2. I will shortly present why.
```{r gauss_basis}
gauss_basis_gen <- function(data_frame, s_sq = 0.02) {
#  x <- data_frame$x
  result <- data_frame
  counter <- 1
  for (i in seq(-1, 1, 0.25)) {
    result <- result %>% 
      mutate(gauss_basis(x, mu = i, s_sq = s_sq))
    colnames(result)[dim(result)[2]] <- paste('gauss', counter, sep = "_")
    counter <- counter + 1
  }
  return(result)
}
```

Now, we generate data using presented functions and plot them.
```{r graph}
my_values <- gauss_basis_gen(data_gen(100), s_sq = 0.02)

my_values %>% 
  melt(id.vars = c("x", "target"), variable.name = "gauss_fun") %>% 
  ggplot(aes(x = x, y = value)) + 
  geom_line(aes(colour = gauss_fun), size = 1) + 
  geom_point(aes(y = target), alpha = 0.2, color = "indianred") + 
  scale_color_viridis(discrete = TRUE, option = "inferno") + 
  theme(plot.subtitle = element_text(vjust = 1), 
    plot.caption = element_text(vjust = 1), 
    panel.grid.major = element_line(colour = "gray94"), 
    panel.grid.minor = element_line(linetype = "blank"), 
    panel.background = element_rect(fill = NA))
```


We can observe that selected $s^2 = 0.2$ nicely fits on our  target data.



## Exercise 2 - Regularized Least squares
_Estimate the parameters using regularized least squares for sample sizes: 20,40,…,100. provide an overview table with point estimates for the parameters. (2 points)_

In this section we first create function, which directly calculates RLS. This is done analytically.
Function also checks, whether input is matrix.
```{r regularized LS}
RLS <- function(m, target, lambda) {
  assert_that(is.matrix(m))
  assert_that(is.matrix(target))
  
  w_RLS <- (solve(lambda * diag(dim(m)[2]) + t(m) %*% m) %*% t(m)) %*% target
  return(w_RLS)
}

```



Now we run for loop, each for different amount of data. There is trick used that I multiply i by twenty, and therefore create different amount of data using data_gen function.

```{r}

w_RLS_result <- matrix(ncol = 5, nrow = 9)
colnames(w_RLS_result) <- c(paste("n", (1:5) * 20, sep = "_"))

for (i in 1:5) {
  values_gb <- data_gen(i * 20) %>% 
    gauss_basis_gen()
  
  phi_mat <- values_gb %>% 
    select(starts_with("gauss")) %>% 
    as.matrix()
  
  t_mat <- as.matrix(values_gb$target)
  
  w_RLS_result[, i] <- RLS(m = phi_mat, 
                                   target = t_mat, 
                                   lambda = 0.2)
}


w_RLS_result
```


##Exercise 3

_Estimate the parameters using bayesian linear regression assuming that the parameters have multivariate normal prior: 20, 40,…,100. Here, you can preselect the $\beta$ and $\alpha$. Provide an overview table with point estimates for the parameters. (2 points)_

In this part I use two different methods. I use Bayesian Linear Regression with and without updating
```{r}
BLR <- function(m, target, alpha, beta) {
  assert_that(is.matrix(m))
  assert_that(is.matrix(target))
  
  m_0 <- rep(0, 9)
  S_0 <- alpha * diag(1, 9)
  
  
  S_N <- solve(solve(S_0) + beta * t(m) %*% m)
  m_N <- S_N %*% (solve(S_0) %*% m_0 + beta * t(m) %*% target)
  
  
  return(m_N)
}
```



```{r}
BLR_updating <- function(m, target, alpha, beta) {
  assert_that(is.matrix(m))
  assert_that(is.matrix(target))
  #setting prior
  m_0 <- as.matrix(rep(0, 9))
  S_0 <- alpha * diag(1, 9)
  # for proper working, m_N needs to be defined beforehand. Let's set it the same.
  m_N <- m_0
  S_N <- S_0
  
  
  for (i in 1:dim(m)[1]) {
    m_0 <- m_N # old posterior is new prior.
    S_0 <- S_N
    phi <- t(as.matrix(m[i, ]))
    S_N <- solve(solve(S_0) + (beta * t(phi) %*% phi))
    m_N <- S_N %*% (solve(S_0) %*% m_0 + (beta * t(phi) %*% target[i]))
  }
  
  return(m_N)
  
}
```





```{r BLR results}
w_BLR_result <- matrix(ncol = 5, nrow = 9)
colnames(w_BLR_result) <- c(paste("n", (1:5) * 20, sep = "_"))

w_BLRupd_result <- matrix(ncol = 5, nrow = 9)
colnames(w_BLRupd_result) <- c(paste("n", (1:5) * 20, sep = "_"))



for (i in 1:5) {
  values_gb <- data_gen(i * 20) %>% 
    gauss_basis_gen()
  
  phi_mat <- values_gb %>% 
    select(starts_with("gauss")) %>% 
    as.matrix()
  
  t_mat <- as.matrix(values_gb$target)
  
  w_BLR_result[, i] <- BLR(m = phi_mat, 
                                   target = t_mat, 
                                   alpha = 0.1, beta = 25)
  w_BLRupd_result[, i] <- BLR_updating(m = phi_mat, 
                                   target = t_mat, 
                                   alpha = 0.1, beta = 25)
}

w_BLR_result

```

```{r}
w_BLRupd_result
```


```{r}
# values_gb <- data_gen(100) %>% 
#     gauss_basis_gen()
#   
# phi_mat <- values_gb %>% 
#     select(starts_with("gauss")) %>% 
#     as.matrix()


values_gb$pred_BLR <- (phi_mat %*% as.matrix(w_BLR_result[,5]))[, 1]
values_gb$pred_BLRupd <- (phi_mat %*% as.matrix(w_BLRupd_result[,5]))[, 1]
values_gb$pred_RLS <- (phi_mat %*% as.matrix(w_RLS_result[,5]))[, 1]

values_gb %>% 
  ggplot() + 
  geom_line(aes(x = x, y = pred_BLR), color = "red") + 
  geom_line(aes(x = x, y = (pred_BLRupd + 0.01)), color = "blue") +
  geom_line(aes(x = x, y = (pred_RLS)), color = "green") +
  geom_line(aes(x = x, y = sin(2*pi*x))) + 
  geom_jitter(aes(x = x, y = target))


```


```{r}
get_w_BLR <- function(my_values, target, alpha, beta){
  phi_mat <- my_values %>% 
    select(3:11) %>% 
    as.matrix()
  t_mat <- as.matrix(target)
  return(BLR_updating(m = phi_mat, target = t_mat, alpha = alpha, beta = beta))
}


w_BLR_result <- matrix(ncol = 5, nrow = 9)
colnames(w_BLR_result) <- c(paste("n", (1:5) * 20, sep = "_"))

for (i in 1:5) {
  values <- data_gen(i * 20)
  target <- values$target
  x <- values$x
  w_BLR_result[, i] <- get_w_BLR(gauss_basis_gen(values), target = target, alpha = 0.1, beta = 2)
}


w_BLR_result
```



```{r}
values <- data_gen(100)
val_res <- values %>% gauss_basis_gen(s_sq = 0) %>% mutate(s_sq = 0)

# create dataset of values for different S_sq
for (i in 1:25) {
  val_temp <- values %>% 
    gauss_basis_gen(s_sq = 0.01 * i) %>% 
    mutate(s_sq = 0.01 * i)
  val_res <- rbind(val_res, val_temp)
}

# finding coefficients and then predictions for every set of data.
val_res_2 <- values %>% mutate(preds = 0) %>% mutate(s_sq = 0)
for (i in unique(val_res$s_sq)) {
  val_temp <- val_res %>% filter(s_sq == i)
  phi_mat <- val_temp %>% 
    select(starts_with("gauss")) %>% 
    as.matrix()
  t_mat <- as.matrix(val_temp$target)
  coefs <- RLS(m = phi_mat, 
              target = t_mat, 
              lambda = 0.2)
  val_temp_2 <- cbind(values, phi_mat %*% as.matrix(coefs))
  val_temp_2 <- cbind(val_temp_2, rep(i, 100))
  colnames(val_temp_2) <- names(val_res_2)
  val_res_2 <- rbind(val_res_2, val_temp_2)
}
  
  

val_res_2 <- val_res_2 %>% filter(s_sq != 0)

# plotting !
abc <- ggplot(val_res_2, aes(x = x, y = preds)) +
  geom_point(aes(frame = s_sq)) + 
  geom_line(aes(x = x, y = sin(2*pi*x)))
abc <- ggplotly(abc)                

abc
```

